Summary of the things that I learned

### 2024-11-08 200-category-theory.tex
Add chapter introductions

I will assume a familiarity with the category-theoretical concepts presented in \autocite{CT4P}. These include categories, functors, isomorphisms, natural transformations, adjunctions, equivalences and limits.

@@

This gives us what we need to translate the classical equivalence between the two notions of Karoubi envelope to type theory:
\begin{lemma}[\coqident{AlgebraicTheories.FundamentalTheorem.CommonUtilities.KaroubiEnvelope}{karoubi_envelope_is_retract}]\label{lem:karoubi-equivalence}
  We have an adjoint equivalence $ \overline \C \xrightarrow \sim \tilde \C $.
\end{lemma}
\begin{proof}
  As shown in Remark \ref{rem:retract-coequalizer}, an object $ (X, f) : \overline \C $ is the equalizer of $ \id X $ and $ f $. Therefore, we send it to the equalizer
  \begin{center}
    \begin{tikzcd}
      F \arrow[r, "s"] & \yo(X) \arrow[r, shift left, "\id{\yo(X)}"] \arrow[r, shift right, "\yo(f)"'] & \yo(X)
    \end{tikzcd}
  \end{center}
  Note that for $ \yo(f) $, we have $ \yo(f) \cdot \yo(f) = \yo(f) \cdot \id{\yo(X)} $, so the universal property of the equalizer gives a morphism $ r: P\C(\yo(X), F) $ such that $ r \cdot s = \yo(f) $. Using this same universal property, we can show that $ s \cdot r = \id F $.
  We send a morphism $ g: \overline \C((X_1, f_1), (X_2, f_2)) $ to
  \[ s_1 \cdot \yo(g) \cdot r_2 : P\C(F_1, F_2). \]
  Note that this is an equivalence on the morphisms: For any morphism $ \bar g : P\C(F_1, F_2) $, we have
  \[ r_1 \cdot \bar g \cdot s_2 : P\C(\yo(X_1), \yo(X_2)). \]
  By the fully faithfulness of the Yoneda lemma, this corresponds to a morphism $ g : \C(X_1, X_2) $ and we can show that these two maps between $ \C((X_1, f_1), (X_2, f_2)) $ and $ \overline \C(F_1, F_2) $ are inverses of each other.
  \begin{center}
    \begin{tikzcd}
      F_1 \arrow[r, shift left, "s_1"] \arrow[d, "\bar g"] & \yo(X_1) \arrow[l, shift left, "r_1"] \arrow[r, shift left, "\id{\yo(X_1)}"] \arrow[r, shift right, "\yo(f_1)"'] \arrow[d, "\yo(g)"] & \yo(X_1)\\
      F_2 \arrow[r, shift left, "s_2"] & \yo(X_2) \arrow[l, shift left, "r_2"] \arrow[r, shift left, "\id{\yo(X_2)}"] \arrow[r, shift right, "\yo(f_2)"'] & \yo(X_2)\\
    \end{tikzcd}
  \end{center}

@@

\begin{remark}
  As we saw in this section, there are multiple candidates for the `Karoubi envelope'. Two of these candidates are equivalent to each other, but the third is not. Therefore, we need to be careful which definition we choose, because this choice has consequences. On one hand, $ \hat \C $ is univalent, but very abstract. In this category, there are classical constructions that we cannot do, because for every object, we only have mere existence of an idempotent morphism, and because the objects of $ \hat \C $ do not form a set, we cannot use the axiom of choice to pick an idempotent morphism. On the other hand, $ \overline \C $ and $ \tilde \C $ are very elementary and concrete, which sometimes helps when doing constructions. However, they are not univalent, which makes working with them complicated in a different way.
\end{remark}

>>>>

This chapter will introduce most of the category theory that is needed to understand the rest of the thesis. The chapter assumes familiarity with the category theoretical concepts presented in \autocite{CT4P}. These include categories, functors, isomorphisms, natural transformations, adjunctions, equivalences and limits.

It is probably wise to skim this chapter upon first reading, and periodically come back to it to better understand the material in the other chapters.

@@

This gives us what we need to translate the classical equivalence between the two notions of Karoubi envelope to type theory:
\begin{lemma}[\coqident{AlgebraicTheories.FundamentalTheorem.CommonUtilities.KaroubiEnvelope}{karoubi_equivalence}]\label{lem:karoubi-equivalence}
  We have an adjoint equivalence $ \overline \C \xrightarrow \sim \tilde \C $.
\end{lemma}
\begin{proof}
  As shown in Remark \ref{rem:retract-coequalizer}, an object $ (X, f) : \overline \C $ is the equalizer of $ \id X $ and $ f $. Therefore, we send it to the equalizer
  \begin{center}
    \begin{tikzcd}
      F \arrow[r, "s"] & \yo(X) \arrow[r, shift left, "\id{\yo(X)}"] \arrow[r, shift right, "\yo(f)"'] & \yo(X)
    \end{tikzcd}
  \end{center}
  Note that for $ \yo(f) $, we have $ \yo(f) \cdot \yo(f) = \yo(f) \cdot \id{\yo(X)} $, so the universal property of the equalizer gives a morphism $ r: P\C(\yo(X), F) $ such that $ r \cdot s = \yo(f) $. Using this same universal property, we can show that $ s \cdot r = \id F $.
  We send a morphism $ g: \overline \C((X_1, f_1), (X_2, f_2)) $ to
  \[ s_1 \cdot \yo(g) \cdot r_2 : P\C(F_1, F_2). \]
  Note that this is an equivalence on the morphisms: For any morphism $ \bar g : P\C(F_1, F_2) $, we have
  \[ r_1 \cdot \bar g \cdot s_2 : P\C(\yo(X_1), \yo(X_2)). \]
  By the fully faithfulness of the Yoneda lemma, this corresponds to a morphism $ g : \C(X_1, X_2) $ and we can show that these two maps between $ \C((X_1, f_1), (X_2, f_2)) $ and $ \overline \C(F_1, F_2) $ are inverses of each other.
  \begin{center}
    \begin{tikzcd}
      F_1 \arrow[r, shift left, "s_1"] \arrow[d, "\bar g"] & \yo(X_1) \arrow[l, shift left, "r_1"] \arrow[r, shift left, "\id{\yo(X_1)}"] \arrow[r, shift right, "\yo(f_1)"'] \arrow[d, "\yo(g)"] & \yo(X_1)\\
      F_2 \arrow[r, shift left, "s_2"] & \yo(X_2) \arrow[l, shift left, "r_2"] \arrow[r, shift left, "\id{\yo(X_2)}"] \arrow[r, shift right, "\yo(f_2)"'] & \yo(X_2)\\
    \end{tikzcd}
  \end{center}

@@

\begin{remark}\label{rem:karoubi-candidate-choice}
  As we saw in this section, there are multiple candidates for the `Karoubi envelope'. Two of these candidates are equivalent to each other, but the third is not. Therefore, we need to be careful which definition we choose, because this choice has consequences. On one hand, $ \hat \C $ is univalent, but very abstract. In this category, there are classical constructions that we cannot do, because for every object, we only have mere existence of an idempotent morphism, and because the objects of $ \hat \C $ do not form a set, we cannot use the axiom of choice to pick an idempotent morphism. On the other hand, $ \overline \C $ and $ \tilde \C $ are very elementary and concrete, which sometimes helps when doing constructions. However, they are not univalent, which makes working with them complicated in a different way.
\end{remark}

### 2024-11-08 300-univalent-foundations.tex
Add chapter introductions

\section{Dependent Type Theory}
Univalent foundations takes place in a framework of type theory. In this section, we will quickly introduce some topics that we will need in subsequent sections.

@@

\section{The Univalence Axiom}
Now, even for a basic category, like the category of types, it seems impossible to prove that the univalence principle holds. However, this is no surprise: It turns out that there is a model in set theory for Martin-Löf type theory with univalence \autocite{simplicial-set-model}, but also for Martin-Löf type theory with proof irrelevance \autocite{proof-irrelevant-model}. In the latter, the type $ x = y $ has at most one element for all $ x, y : T $ and all types $ T $, which contradicts univalence. Therefore, univalence is independent of the axioms of Martin-Löf type theory: both univalence or its negation cannot be proved, but either can be assumed as an axiom.

@@

\section{Truncation and (Mere) Existence}
As mentioned before, if we want to do mathematics in a dependent type theory, we can `encode' propositions as types. The elements of the type correspond to the proofs that the proposition is true, see for example the identity types. However, we need to be careful here about the distinction between types in general and mere propositions:

@@

\section{Equality and Homotopy}
Univalent Foundations is often mentioned together with Homotopy Type Theory, because they are related but distinct concepts. Therefore, we will mention it here.

@@

\section{Transport}

>>>>

Since this thesis is, among other things, a case study in translating mathematics from set theory to univalent foundations, this chapter will give an introduction to univalent foundations and introduce some of the relevant concepts. It will also point out some issues that can arise in such a translation, like the difference between existence and mere existence (Section \ref{sec:truncation-mere-existence}).

First, we will introduce dependent type theory (Section \ref{sec:dependent-type-theory}). After this, we will talk about the univalence principle (Section \ref{sec:univalence-principle}), and introduce the univalence axiom (Section \ref{sec:univalence-axiom}). Then we will discuss how propositions and sets show up in univalent foundations (Section \ref{sec:homotopy-props-sets}) and introduce the related notions of propositional truncation and `mere existence' (Section \ref{sec:truncation-mere-existence}). After this, we will discuss the relation between equality and homotopy, which is a central idea in homotpy type theory (Section \ref{sec:equality-homotopy}), and lastly, we will talk about transport over equalities (Section \ref{sec:transports}).

\section{Dependent Type Theory}\label{sec:dependent-type-theory}
Univalent foundations takes place in a framework of type theory. In this section, we will quickly introduce some topics that we will need in subsequent sections.

@@

\section{The Univalence Axiom}\label{sec:univalence-axiom}
Now, even for a basic category, like the category of types, it seems impossible to prove that the univalence principle holds. However, this is no surprise: It turns out that there is a model in set theory for Martin-Löf type theory with univalence \autocite{simplicial-set-model}, but also for Martin-Löf type theory with proof irrelevance \autocite{proof-irrelevant-model}. In the latter, the type $ x = y $ has at most one element for all $ x, y : T $ and all types $ T $, which contradicts univalence. Therefore, univalence is independent of the axioms of Martin-Löf type theory: both univalence or its negation cannot be proved, but either can be assumed as an axiom.

@@

\section{Truncation and (Mere) Existence}\label{sec:truncation-mere-existence}
As mentioned before, if we want to do mathematics in a dependent type theory, we can `encode' propositions as types. The elements of the type correspond to the proofs that the proposition is true, see for example the identity types. However, we need to be careful here about the distinction between types in general and mere propositions:

@@

\section{Equality and Homotopy}\label{sec:equality-homotopy}
Univalent Foundations is often mentioned together with Homotopy Type Theory, because they are related but distinct concepts. Therefore, we will mention it here.

@@

\section{Transport}\label{sec:transports}

### 2024-11-08 400-algebraic-structures.tex
Add chapter introductions

@@

\section{Algebras}

@@

\section{Presheaves}

@@

\section{\texorpdfstring{$ \lambda $-}{lambda-}theories}

@@

\section{Examples}

>>>>

In his paper, Hyland uses concepts from universal algebra to study the $ \lambda $-calculus. This chapter introduces these concepts: it consecutively introduces algebraic theories (Section \ref{sec:algebraic-theories}), algebras (Section \ref{sec:algebras}), presheaves (Section \ref{sec:presheaves}) and $ \lambda $-theories (Section \ref{sec:lambda-theories}). The last part of this chapter, Section \ref{sec:algebraic-theory-examples}, is devoted to examples, so it may be useful to skip back and forth a couple of times to get a better understanding of what the definitions are about.

@@

\section{Algebras}\label{sec:algebras}

@@

\section{Presheaves}\label{sec:presheaves}

@@

\section{\texorpdfstring{$ \lambda $-}{lambda-}theories}\label{sec:lambda-theories}

@@

\section{Examples}\label{sec:algebraic-theory-examples}

### 2024-11-08 500-previous-work.tex
Add chapter introductions

% \begin{tabular}{|c|c|c|}\hline
%   $ K $ & $ \lambda xy.x $ & constant \\\hline
%   $ S $ & $ \lambda xyz.xz(yz) $ & \\\hline
%   $ I $ & $ \lambda x.x = SKK $ & identity \\\hline
%   $ Y $ & $ \lambda f.(\lambda x.f (xx))(\lambda x.f (xx)) $ & ﬁxpoint \\\hline
%   $ \bot $ & $ (\lambda x.xx)(\lambda x.xx) = (SII)(SII) = YI $ & bottom \\\hline
%   $ P $ & $ \lambda f gx.g(f x) = S(K(S(S(KS)K)))K $ & composition \\\hline
%   $ Q $ & $ \lambda f gax.ga(f ax) $ & ﬁbred composition \\\hline
%   $ \langle \rangle $ & $ \lambda xyz.zxy $ & pairing \\\hline
%   $ 0 $ & $ \lambda xy.x = K $ & left projection \\\hline
%   $ 1 $ & $ \lambda xy.y = KI $ & right projection \\\hline
% \end{tabular}

@@

\section{Scott's Representation Theorem}
The correspondence in Section \ref{sec:lambek-correspondence} between simply-typed $ \lambda $-calculi and cartesian closed categories raises a question whether such a correspondence also exists for untyped $ \lambda $-calculi. Definition \ref{def:endomorphism-theory} shows that in fairly general circumstances we can take one object $ c $ in a category $ C $ and consider the morphisms $ t: C(c^n, c) $ as terms in an untyped $ \lambda $-calculus. Hyland calls this the `endomorphism theory' of $ c $.

@@

\section{The Taylor Fibration}

@@

\section{The \texorpdfstring{$ \lambda $}{lambda}-calculus is Algebraic}
As we mentioned in Remark \ref{rem:lambda-theory-or-algebra}, using tools from universal algebra, there are two approaches to study $ \lambda $-calculus-like structures. One can either study $ \lambda $-theories or algebras for the algebraic theory $ \Lambda $. We will see that Hyland shows that these are equivalent.

>>>>

Hyland, like every scientist, stands on the shoulder of giants. As we saw in the last chapter, he borrows concepts from universal algebra to study the $ \lambda $-calculus. Furthermore, two main theorems in his paper have been proven by others before him, although in a different way. This chapter strives to improve the understanding of Hyland's work by expositing the work of those who came before.

To understand what categorical semantics for the untyped $ \lambda $-calculus is about, we first briefly take a look at categorical semantics for the simply typed $ \lambda $-calculus (Section \ref{sec:lambek-correspondence}). After this, we provide Scott's result about categorical semantics for the untyped $ \lambda $-calculus (Section \ref{sec:scott}), for which we first study the `category of retracts' (Section \ref{sec:retracts-category}). We then discuss Paul Taylor's result about the structure of this category of retracts (Section \ref{sec:taylor}), which says something about its internal logic. Lastly, we briefly discuss a result of Peter Selinger (Section \ref{sec:selinger}), which seems to have inspired Hyland's fundamental theorem of the $ \lambda $-calculus.

@@

\section{Scott's Representation Theorem}\label{sec:scott}
The correspondence in Section \ref{sec:lambek-correspondence} between simply-typed $ \lambda $-calculi and cartesian closed categories raises a question whether such a correspondence also exists for untyped $ \lambda $-calculi. Definition \ref{def:endomorphism-theory} shows that in fairly general circumstances we can take one object $ c $ in a category $ C $ and consider the morphisms $ t: C(c^n, c) $ as terms in an untyped $ \lambda $-calculus. Hyland calls this the `endomorphism theory' of $ c $.

@@

\section{The Taylor Fibration}\label{sec:taylor}

@@

\section{The \texorpdfstring{$ \lambda $}{lambda}-calculus is Algebraic}\label{sec:selinger}
As we mentioned in Remark \ref{rem:lambda-theory-or-algebra}, using tools from universal algebra, there are two approaches to study $ \lambda $-calculus-like structures. One can either study $ \lambda $-theories or algebras for the algebraic theory $ \Lambda $. We will see that Hyland shows that these are equivalent.

### 2024-11-08 600-paper.tex
Add chapter introductions

\section{Scott's Representation Theorem}

@@

\section{Relations Between the Categories}

@@

\section{Hyland's Proof}

@@

\section{The Theory of Extensions}

>>>>

In this chapter, we will cover the main results from Hyland's paper. We work out the proofs in more detail, and point out where subtleties arise when translating them from set theory to univalent foundations.

The chapter starts with a new proof for Scott's representation theorem (Section \ref{sec:hyland-scott}). The sections after this deal with a lot of different categories and Section \ref{sec:hyland-categories} discusses the relations between these categories. Section \ref{sec:relatively-cartesian-closed} gives Hyland's proof of the result by Paul Taylor that was covered in the previous chapter. Here, the bifurcation of the Karoubi envelope in univalent foundations, as mentioned in Remark \ref{rem:karoubi-candidate-choice}, means that Hyland's result is about a different category than Taylor's result (Remark \ref{rem:difference-Taylor-Hyland}).

In the last part of this chapter, we cover Hyland's fundamental theorem about the equivalence between $ \lambda $-theories and $ \Lambda $-algebras. There are multiple ways to construct a $ \lambda $-theory from a $ \Lambda $-algebra, and these give rise to slightly different proofs. We will consecutively cover a very elementary version (Section \ref{sec:elementary-fundamental-theorem}), Hyland's first version, which uses the endomorphism theory (Section \ref{sec:hyland-fundamental-theorem}), and Hyland's second version, which uses the theory of extensions (Section \ref{sec:theory-of-extentions}).

\section{Scott's Representation Theorem}\label{sec:hyland-scott}

@@

\section{Relations Between the Categories}\label{sec:hyland-categories}

@@

\section{Hyland's Proof}\label{sec:hyland-fundamental-theorem}

@@

\section{The Theory of Extensions}\label{sec:theory-of-extentions}

### 2024-11-08 910-alternative-definitions.tex
Add chapter introductions

\chapter{Alternative definitions}\label{ch:alternative-definitions}
The literature, there are many different but equivalent definitions, carrying many different names, for the objects that called `algebraic theories' in Section \ref{sec:algebraic-theories}. Also, many of these different names are attached to differently defined objects in different sources. This section will showcase some of the various definitions.

In this appendix, we will denote the finite set $ \llbracket n \rrbracket = \{ 1, \dots, n \} $.

@@

\section{Lawvere Theory}
\begin{definition}
  An \iindex{algebraic theory} as presented in \autocite{algebraic-theories-2010} is a small category with finite products.
\end{definition}
\begin{definition}
  An \iindex{algebra} for an algebraic theory $ T $ is a finite-product-preserving functor $ T \to \SET $.
\end{definition}

>>>>

\chapter{Alternative definitions}\label{ch:alternative-definitions}
In the literature, there are many different but equivalent definitions, carrying many different names, for the objects that are called `algebraic theories' in Section \ref{sec:algebraic-theories}. To make matters even more confusing, the same name can refer to different things in different parts of the literature. This appendix will showcase various definitions and give them a name. Of particular interest is the Lawvere theory (Section \ref{sec:lawvere-theory}), which is also used in Hyland's paper.

In this appendix, we will denote the finite set $ \{ 1, \dots, n \} $ as $ \llbracket n \rrbracket $.

@@

\section{Lawvere Theory}\label{sec:lawvere-theory}
\begin{definition}
  An \iindex{algebraic theory} as presented in \autocite{algebraic-theories-2010} is a small category with finite products.
\end{definition}
\begin{definition}
  An \iindex{algebra} for an algebraic theory $ T $ is a finite-product-preserving functor $ T \to \SET $.
\end{definition}

### 2024-11-08 920-weak-cartesian-closed.tex
Add chapter introductions

\chapter{Weak Cartesian Closed Categories}\label{ch:weak-cartesian-closed-categories}
In Definition \ref{def:endomorphism-theory}, we defined the endomorphism $ \lambda $-theory of a reflexive object in a cartesian closed category. However, as it turns out, we do not need all of the properties of a cartesian closed category. It suffices to consider a \textit{weak} cartesian closed category. In fact, even a `semi cartesian closed category' will suffice, but in this section, we will restrict ourselves to a weak cartesian closed category, as this notion suits our purposes well enough.

>>>>

\chapter{Weak Cartesian Closed Categories}\label{ch:weak-cartesian-closed-categories}
In Definition \ref{def:endomorphism-theory}, we defined the endomorphism $ \lambda $-theory of a reflexive object in a cartesian closed category. However, as it turns out, we do not need all of the properties of a cartesian closed category. It suffices to consider a \textit{weak} cartesian closed category, which we will discuss in this appendix. In fact, even a `semi cartesian closed category' will suffice, but in this appendix, we will restrict ourselves to a weak cartesian closed category, as this notion suits our purposes well enough.
